Finiteness in Cubical Type Theory

### 2020-05-27 paper.tex
some new intro stuff

\section{Introduction}
\subsection{Foreword}
In constructive mathematics we are often preoccupied with \emph{why} something
is true.
Take finiteness, for example.
There are a handful of ways to demonstrate some type is finite: we could provide
a surjection to it from another finite type; we could show that any collection
of its elements larger than some bound contains duplicates; or we could show
that any stream of its elements contain duplicates.

>>>>

\section{Introduction}
\subsection{Foreword}
Proofs in constructive mathematics are often more substantial than their
classical counterparts.
``Substantial'' here doesn't mean the literal length of the proofs (although the
double meaning is not lost on anyone who has had to struggle with a modern
constructive proof assistant), but rather refers to their \emph{content}: a
constructive proof contains more \emph{stuff} than a classical one.

If we were to prove that a set was finite in classical mathematics, we might
show that it is in bijection with another finite set, say a finite prefix of the
natural numbers.
In semi-formal notation, to say that a set \(A\) is finite we would show the
following:
\begin{equation} \label{example-finite}
  \mathbf{Finite}(A) \coloneqq \exists (n : \mathbb{N}) . \; A \Leftrightarrow \left\{ m \mid m < n\right\}
\end{equation}
Where \(A \Leftrightarrow B\) means ``a bijection between \(A\) and \(B\)''.

Classically, a proof of equation~\ref{example-finite} is nothing more than
\emph{evidence} for the proposition of equation~\ref{example-finite}.
A constructive proof of \ref{example-finite}, however, could include much more:
\begin{enumerate}
  \item A number \(n\) representing the cardinality of \(A\).
  \item A function from \(A\) to \(\left\{ m \mid m < n \right\}\).
  \item A function from \(\left\{ m \mid m < n \right\}\) to \(A\).
\end{enumerate}
In other words, it makes sense constructively to talk about a function of type
\(\mathbf{Finite}(A) \rightarrow \mathbb{N}\), which takes a \emph{proof} of the
finiteness of \(A\) and returns \(A\)'s cardinality.

Such functions are perhaps more concrete than this description suggests:
aside from being interesting in its own right, much of the attention given to
constructive mathematics comes from its perceived association with
computers (themselves a burgeoning topic of study, bound to catch on some day). 
Indeed many constructive proofs are in fact computer programs in disguise (and
vice-versa \cite{wadlerPropositionsTypes2015}): the functions referred to above
correspond to honest-to-goodness functions in a programming language (Agda
\cite{vezzosiCubicalAgdaDependently2019} in our case) which can be run on a
computer to produce real output.

All of this added information in our proofs is interesting and useful, yes, but
it also necessitates some extra caution.
We will see cases where we have provided \emph{too much} information in a
constructive proof, and we will need to hide it somehow.
For instance, in order to construct a category of finite sets we will need to
describe the objects of that category: constructively this would be ``pairs of
sets along with proofs that they are finite''.
This gives us the wrong category, however: for a type \(A\) with cardinality
\(n\), there are \(n!\) proofs that \(A\) is finite.
This means that we will have \(n!\) objects in the category for the set \(A\),
where we only want 1.
The solution is to hide some aspect of the proof: this is accomplished with
\emph{truncation}, which we will see in \todo{ref}.

For all of its benefits, constructive mathematics (especially dependently-typed
constructive mathematics on a computer) still suffers from many painful
drawbacks.
As already hinted at, proofs in a constructive setting can tend to be verbose,
especially when the arbiter of correctness is as strict as a computer, which
neither understands nor trusts the time-honoured technique of ``the other cases
are obviously true''.
While the goal of using a computer to automate the boring parts of a proof still
seems attainable, too often getting a computer involved in a proof means adding
thousands of lines of tedium to satisfy its lack of intuition.

In this paper we are interested in two things: systematic generation of data,
and constructive finiteness.

First, let's explain what we mean by ``systematic generation of data''.
Consider QuickCheck \cite{claessenQuickCheckLightweightTool2011}: this
well-known Haskell library allows us to quickly and easily test
\ExecuteMetaData[haskell/PropertyTests.tex]{sort-stable}

In constructive mathematics we are often preoccupied with \emph{why} something
is true.
Take finiteness, for example.
There are a handful of ways to demonstrate some type is finite: we could provide
a surjection to it from another finite type; we could show that any collection
of its elements larger than some bound contains duplicates; or we could show
that any stream of its elements contain duplicates.
